Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2016-09-20 document/ms.tex
Relates to Issue #47-- feedback from GJH, new h2o jump feature figure added, some rearranging.

\author{Michael A. Gully-Santiago,\altaffilmark{1} Gregory J. Herczeg,\altaffilmark{1} Ian Czekala,\altaffilmark{2} Garrett Somers,\altaffilmark{3} Konstantin Grankin,\altaffilmark{4} Kevin Covey,\altaffilmark{5} J.F. Donati,\altaffilmark{6} Sylvia Alencar,\altaffilmark{7}, Gaitee Hussein,\altaffilmark{8} Ben Shappee,\altaffilmark{9} Greg Mace,\altaffilmark{10} Jae-Joon Lee\altaffilmark{11} T.~W.-S.~Holoien,\altaffilmark{12,13} Jessy Jose\altaffilmark{1} Jimmy Liu\altaffilmark{15}}

@@

\subsection{Absolute upper limit of starspot temperature}

Consider relaxing assumption 1 from above to better match reality: a starspot is not at absolute zero temperature, it has a finite temperature and therefore emits a finite amount of radiation.  Increasing the temperature a little means that the starspot size has to increase to cause the same flux difference.  The starspot cannot occupy more than 100\% of the projected stellar hemisphere, so there is an upper limit for the temperature of the starspot that yields a given $\Delta V$.  The starspot temperature upper limit for $\Delta V=0.5$ is 3650 K.  In the 2004 season, the starspot temperature upper limit was 3410 K.

In reality, the star will have some patches of starspots that always face toward the observer, like polar spots, bands of spots, multiple finitie-sized patches of starspots, or homogenuously distributed small spots.  Dark spots on extreme latitudes on inclined stars might only pass over the limb of the stellar disk, reducing the diminution $V-$band flux for a given projected surface area.  All of these effect mean that real starspot filling factors are likely to be larger than the filling factor lower limit, and cooler than the temperature upper limit.

@@

\begin{figure*}
 \centering
  \includegraphics[width=0.45\textwidth]{figures/v_v-r_Grankin} 
 \begin{tabular}{ll}
 \includegraphics[trim=2.6cm 13.0cm 2.8cm 3.2cm, clip=true, width=0.45\textwidth]{figures/bv_pec13} 
    &
   \includegraphics[trim=2.6cm 13.0cm 2.8cm 2.2cm, clip=true, width=0.45\textwidth]{figures/vr_pec13}
    \end{tabular}
\caption{The observed optical colors of LkCa 4 from \citet{grankin08}, compared with predictions.  
\emph{Top}: Grey points are all \citet{grankin08} photometric data, while black points are the subset of observations from the 1999 season.  The purple line represents a two component photometric model  described by \citet{grankin08}, comprising a photospheric temperature of 4040K (for K7), and a temperature of spotted area of 3030K (for an M5). In this model, the spot coverage of the observed hemisphere varies from $\sim36\%$ at maximum brightness, and $\sim82\%$ at minimum brightness.
\emph{Bottom:}  The model is constructed instead by converting $V$-band brightness to a cool spot filling factor and subsequently calculating colors from main sequence colors and bolometric corrections of \citet{pecaut13} (purple lines).}
\label{fig:colors}
\end{figure*}

@@

An accurate extinction estimate from the full SED is challenging because of a strong dependence on the areal distribution of temperatures in the photosphere, which in our work is simplified to a two temperature fit.  However, isolating specific colors constrains the possible range of extinctions.  
The $B-V$ color is stable and not affected by cooler components, and may therefore be compared with the expected colors for the hotter (4100 K) temperature.  
The average $B-V=1.41$ mag color from \citet{grankin08} has a color excess $E(B-V)=0.09$ mag relative to the empirical color expected for a main sequence star and $E(B-V)=0.20$ mag for colors of pre-main sequence stars, based on the temperature-color tables of \citep{pecaut13}.  The main sequence colors are probably most 
appropriate in this analysis because they are more free of spots.  
The $B-V$ color 
excess leads to an extinction $A_V=0.28\pm0.15$ mag, or $0.62\pm0.15$ mag relative to 
pre-main sequence colors.   The listed uncertainties are calculated from extinction estimates relative to colors of 4000 and 4200 K photospheres.
The $V-R_J$ color excess leads to $A_V=0.48$ mag, which should be less affected by gravity (based on $V-I_C$ colors in Pecaut et al.) but is likely to be overestimated because the cool component contributes to the $R_J$-band emission, as seen in Fig.~\ref{fig:colors}.
In the near-IR, the estimated $V=12.61$ mag at the time of the 2MASS observations leads to a template (main sequence) color 
$J-K_s=0.87$ mag, with a range of $\sim 0.05$ mag for acceptable solutions with other 
temperatures and covering fractions. The observed $J-K=0.93$ mag leads to $A_V=0.35$ mag, 
with a range of $0.29$ mag to account for the range of other acceptable photospheric solutions.  
The template near-IR color is more robust than optical colors to differences 
in the main sequence/pre-main sequence temperature scales because the pre-main 
sequence $J-K$ color is redder at 4100 K and bluer at 2750 K.  Based on this 
analysis, an $A_V=0.3$ mag is adopted throughout this paper.

@@

The spectrum of LkCa 4 is well reproduced by a cool component of $2750$ K covering 80\% of the surface and a $4100$ K component covering 20\% of the surface.   Scaling this composite spectrum for LkCa 4 to the observed $K$-band photometry yields $R=2.3$ $R_\odot$ and $\log L/L_\odot=0.17$, adopting $A_V=0.3$ mag and a distance of 131 pc for the parallax distance of the nearest Taurus member \citep{torres12}.  The effective temperature of these two components is 3330 K, as calculated from the average surface flux at the median $V$ band magnitude.  Figure~\ref{fig:hrdiag} and Table~4 shows that this effective temperature is much cooler than previous measurements from optical spectra, which are clustered around 4000 K.  The luminosity does not change significantly because the broadband SED yields similar luminosities, independent of temperature.

@@

\subsection{Addressing limitations of assumptions in the spectral inference methodology}

@@

Finally, uncertainties in the \PHOENIX\ synthetic spectra continuum opacity flow down to uncertainties in the absolute value of the fill factor.  Erroneous bulk opacities could be responsible for some of the scatter of fill factor distributions seen across different spectral orders, as in Figures \ref{fig:TwoTempResults} and \ref{fig:specPostageStamp}.  The large number of spectral orders over which the fill factors are averaged offers resilience against these anomalies.

@@

% Fang
\textbf{Insert Fang info here.}

@@

{\it Facilities:} \facility{Smith (IGRINS)}, \facility{AAVSO}, \facility{CFHT (ESPaDOnS)}, \facility{INTEGRAL (OMC)}, \facility{ASAS}, \facility{CrAO:1.25m}, \facility{ARC (TripleSpec)}, \facility{Hale (DBSP)}

>>>>

\author{Michael A. Gully-Santiago,\altaffilmark{1} Gregory J. Herczeg,\altaffilmark{1} Ian Czekala,\altaffilmark{2} Garrett Somers,\altaffilmark{3} Konstantin Grankin,\altaffilmark{4} Kevin Covey,\altaffilmark{5} J.F. Donati,\altaffilmark{6} Sylvia Alencar,\altaffilmark{7} Gaitee Hussein,\altaffilmark{8} Ben Shappee,\altaffilmark{9} Greg Mace,\altaffilmark{10} Jae-Joon Lee\altaffilmark{11} T.~W.-S.~Holoien,\altaffilmark{12,13} Jessy Jose\altaffilmark{1} Jimmy Liu\altaffilmark{15}}

@@

@@

\begin{figure*}
 \centering
 \begin{tabular}{ll}
 \includegraphics[trim=2.6cm 13.0cm 2.8cm 3.2cm, clip=true, width=0.45\textwidth]{figures/bv_pec13} 
    &
   \includegraphics[trim=2.6cm 13.0cm 2.8cm 2.2cm, clip=true, width=0.45\textwidth]{figures/vr_pec13}
    \end{tabular}
\caption{The observed optical colors of LkCa 4 from \citet{grankin08}, compared with predictions.  The model is constructed instead by converting $V$-band brightness to a cool spot filling factor and subsequently calculating colors from main sequence colors and bolometric corrections of \citet{pecaut13} (purple lines).}
\label{fig:colors}
\end{figure*}

@@

An accurate extinction estimate from the full SED is challenging because of a strong dependence on the areal distribution of temperatures in the photosphere, which in our work is simplified to a two temperature fit.  However, isolating specific colors constrains the possible range of extinctions.  
The $B-V$ color is stable and not affected by cooler components, and may therefore be compared with the expected colors for the hotter (4100 K) temperature.  
The average $B-V=1.41$ mag color from \citet{grankin08} has a color excess $E(B-V)=0.09$ mag relative to the empirical color expected for a main sequence star and $E(B-V)=0.20$ mag for colors of pre-main sequence stars, based on the temperature-color tables of \citep{pecaut13}.  The main sequence colors are probably most 
appropriate in this analysis because they are more free of spots.  
The $B-V$ color 
excess leads to an extinction $A_V=0.28\pm0.15$ mag, or $0.62\pm0.15$ mag relative to 
pre-main sequence colors.   The listed uncertainties are calculated from extinction estimates relative to colors of 4000 and 4200 K photospheres.
The $V-R_J$ color excess leads to $A_V=0.48$ mag, which should be less affected by gravity (based on $V-I_C$ colors in Pecaut et al.) but is likely to be overestimated because the cool component contributes to the $R_J$-band emission, as seen in Fig.~\ref{fig:colors}.
In the near-IR, the estimated $V=12.61$ mag at the time of the 2MASS observations leads to a spot-weighted template (main sequence) color 
$J-K_s=0.87$ mag, with a range of $\sim 0.05$ mag for acceptable solutions with other 
temperatures and covering fractions. The observed $J-K=0.93$ mag leads to $A_V=0.35$ mag, 
with a range of $0.29$ mag to account for the range of other acceptable photospheric solutions.  
The template near-IR color is more robust than optical colors to differences 
in the main sequence/pre-main sequence temperature scales because the pre-main 
sequence $J-K$ color is redder at 4100 K and bluer at 2750 K.  Based on this 
analysis, an $A_V=0.3$ mag is adopted throughout this paper.

@@

The spectrum of LkCa 4 is well reproduced by a cool component of $2750$ K covering 80\% of the surface and a $4100$ K component covering 20\% of the surface.   Scaling this composite spectrum for LkCa 4 to the observed $K$-band photometry yields $R=2.3$ $R_\odot$ and $\log L/L_\odot=0.17$, adopting $A_V=0.3$ mag and a distance of $131 \pm 2$ pc$^x$ based on the range of parallaxes of nearby Taurus members measured by \citet{torres12}\footnote{This distance is consistent with the Gaia distance of $130 \pm4$ pc measured to the nearby Taurus members V410 Tau, V773 Tau, and BP Tau \citep{gaia2016dr}.}.  The effective temperature of these two components is 3330 K, as calculated from the average surface flux at the median $V$ band magnitude.  Figure~\ref{fig:hrdiag} and Table~4 shows that this effective temperature is much cooler than previous measurements from optical spectra, which are clustered around 4000 K.  The luminosity does not change significantly because the broadband SED yields similar luminosities, independent of temperature.

@@

\subsection{Limitations of assumptions in the spectral inference methodology}

@@

Uncertainties in the \PHOENIX\ synthetic spectra continuum opacity flow down to uncertainties in the absolute value of the fill factor.  Erroneous bulk opacities could be responsible for some of the scatter of fill factor distributions seen across different spectral orders, as in Figures \ref{fig:TwoTempResults} and \ref{fig:specPostageStamp}.  The large number of spectral orders over which the fill factors are averaged offers resilience against these anomalies.

\begin{figure*}
 \centering
\includegraphics[width=0.45\textwidth]{figures/v_v-r_Grankin} 
\caption{Grey points are all \citet{grankin08} photometric data, while black points are the subset of observations from the 1999 season.  The purple line represents a two component photometric model  described by \citet{grankin08}, comprising a photospheric temperature of 4040K (for K7), and a temperature of spotted area of 3030K (for an M5). In this model, the spot coverage of the observed hemisphere varies from $\sim36\%$ at maximum brightness, and $\sim82\%$ at minimum brightness.}
 \label{fig:grankin_vr}
\end{figure*}

Analyses of optical emission also suggest that a two-temperature model may be an oversimplification.  Figure \ref{fig:grankin_vr} shows that the $V-R$ colors from \citet{grankin} are better reproduced if the cooler component is $\sim 3050$ K, thereby contributing more flux at optical wavelengths.  The ZDI imaging of \citet{donati14} also indicate a complex structure of spots that are hotter than what is interpreted here as the ambient photosphere, while also reproducing the optical photometry.  Despite the need for more complicated emission structures to simultaneously reproduce all observations, the broadband SED, spectral shape, and rotational modulation of spectral indices are reasonably well fit with the two components presented here.  Moreover, the high and low-resolution near-IR spectra require a cool (2700--3000 K) component while the global photometry requires a large covering fraction for this cool component.

The need for the cool component to have a large covering fraction and $<3000$ K is also demonstrated by the size of the H$_2$O band absorption between the $J$ and $H$ bands (see Figure~\ref{fig:h2ojump}).  Assuming that the hotter component is 4100 K, the H$_2$O opacity is only sufficient if $T_{spot}<3000$ K, with a high spot coverage.

\begin{figure*}
 \centering
 %\includegraphics[width=0.65\textwidth]{}
 \includegraphics[trim=2.6cm 13.0cm 2.3cm 2.5cm, clip=true, width=0.65\textwidth]{figures/jbandjump}
 \caption{The constraint on the size and temperature of the cool spot from the H$_2$O absorption band depth between $J$ and $H$-bands.  The contours show the flux ratio between 1.30 and 1.34 $\mu$m (shaded yellow region in inset plot), as produced by the Phoenix models.  The solid and hatched purple lines show the best fit and range of acceptable contours from the APO/TripleSpec spectrum, which was obtained when the cool spot fraction was $\sim 84$\%.  The range of acceptable contour is based on an estimate uncertainty of $\sim 2$\% in observed flux ratio, which is dominated by uncertainty in telluric correction of strong H$_2$O absorption. }
 \label{fig:h2ojump}
\end{figure*}

%\subsection{Absolute upper limit of starspot temperature}

%Consider relaxing assumption 1 from above to better match reality: a starspot is not at absolute zero temperature, it has a finite temperature and therefore emits a finite amount of radiation.  Increasing the temperature a little means that the starspot size has to increase to cause the same flux difference.  The starspot cannot occupy more than 100\% of the projected stellar hemisphere, so there is an upper limit for the temperature of the starspot that yields a given $\Delta V$.  The starspot temperature upper limit for $\Delta V=0.5$ is 3650 K.  In the 2004 season, the starspot temperature upper limit was 3410 K.

%In reality, the star will have some patches of starspots that always face toward the observer, like polar spots, bands of spots, multiple finitie-sized patches of starspots, or homogenuously distributed small spots.  Dark spots on extreme latitudes on inclined stars might only pass over the limb of the stellar disk, reducing the diminution $V-$band flux for a given projected surface area.  All of these effect mean that real starspot filling factors are likely to be larger than the filling factor lower limit, and cooler than the temperature upper limit.

@@

% Fang
%\textbf{Insert Fang info here.}

@@

{\it Facilities:} \facility{Smith (IGRINS)}, \facility{AAVSO}, \facility{CFHT (ESPaDOnS)}, \facility{INTEGRAL (OMC)}, \facility{ASAS}, \facility{CrAO:1.25m}, \facility{ARC (TripleSpec)}, \facility{Hale (DBSP)}, \facility{Gaia}
